Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-05 formally.tex
missing formally.tex

%\documentclass[12pt]{article}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
\makeatletter

\renewcommand\topfraction{0.85}
\renewcommand\bottomfraction{0.85}
\renewcommand\textfraction{0.1}
\renewcommand\floatpagefraction{0.85}

\AtBeginDocument{%
 \abovedisplayskip=3pt
 \abovedisplayshortskip=0pt
 \belowdisplayskip=2pt
 \belowdisplayshortskip=0pt
}

\setlength{\floatsep}{10pt}
\setlength{\textfloatsep}{12pt}

\usepackage[compact]{titlesec}  
\titlespacing*{\section}{0pt}{3pt}{3pt}
\titlespacing*{\subsection}{0pt}{4pt}{3pt}
\titlespacing*{\subsubsection}{0pt}{4pt}{3pt}
\titlespacing*{\paragraph}{0pt}{3pt}{5pt}
\setlength{\skip\footins}{3pt}

\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink

\usepackage{enumitem}

\input{macros}
\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

%\titlebanner{}        % These are ignored unless
\preprintfooter{Draft}   % 'preprint' option specified.

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus}
%\subtitle{Modular Type Constructors}

\authorinfo{\vspace{-2px}}{}{}
%\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
%         {Carnegie Mellon University}
%         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Programs are rich inductive structures, but programmers typically construct and manipulate them indirectly, through flat textual representations. This indirection  comes at a cost: programmers must understand the  subtleties of parsing, and it can require several text editor actions to make a single syntactically and semantically well-defined change. 
During these sequences of edit actions, or when the programmer makes a mistake, programmers and programming tools must contend with malformed text or semantically ill-defined programs.

\emph{Structure editors} promise to alleviate these burdens by exposing only edit actions that  produce sensible changes to the program structure.
Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically ill-defined as they are being constructed.
In this paper, we report on our ongoing efforts to develop Hazelnut, a minimal  structure editor defined in a principled type-theoretic style where all edit actions leave the program in both a syntactically and semantically well-defined state. Uniquely, Hazelnut does not force the programmer to construct the program in a strictly ``outside-in'' fashion.
Formally, Hazelnut is a bidirectionally typed lambda calculus extended with
1) \emph{holes} (which mark subterms that are being  constructed from the inside out);
2) a \emph{focus model}; and 3) a bidirectional \emph{action model} equipped with a useful \emph{action sensibility} theorem.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation

\section{Introduction}\label{sec:introduction}
\input{intro}

\section{Programming in Hazelnut}\label{sec:example}
\input{example}

\section{Hazelnut, Formally}
\label{sec:hazel}
Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

Whenever an action is performed on a well-typed expression, it produces another well-typed expression in a \emph{sensible} manner. More specifically, the action semantics satisfies a crucial \emph{sensibility theorem}, stated in Sec. \ref{sec:actions}.
\end{itemize}

In our overview of the semantics below, we will reproduce only the most interesting rules, and in some cases we will do so ``out of order.'' The appendix (and our Agda formalization, see Sec. \ref{sec:mech}) defines the complete collection of rules in their dependency order.
\subsection{Holes}\label{sec:holes}
\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \tau,\htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & e,\hexp & ::= &
  \hexp : \htau ~\vert~
  x ~\vert~
  \hlam{x}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}

@@

We refer to terms that do not contain subterms of hole form as \emph{complete}. Informally, we will use metavariables $\tau$ and $e$ rather than $\htau$ and $\hexp$ for complete H-types and H-expressions, respectively. Formally, we can derive $\hcomplete{\tau}$ when $\tau$ is a complete H-type, and $\hcomplete{e}$ when $e$ is a complete H-expression. We omit the straightforward definitions of these judgements for concision. The dynamics of Hazelnut, which we need not detail here, is defined only  over complete H-expressions (i.e. we can only ``run'' a complete program, though see Sec. \ref{sec:future}.)

@@

The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}. Their system (which was not bidirectionally typed nor an editor model) also needed to define two rules for function application. In general, when a premise requires that a synthesized type be of a particular form, we need a special case where the synthesized hole type is treated instead as if it were the ``holey-est'' type of that form.\footnote{Alternatively, we might add a rule that allows expressions that synthesize hole type to then non-deterministically synthesize any other type, but maintaining determinism is useful in practice, so we avoid this approach.}

\end{subequations}
\subsection{Focus Model}\label{sec:cursors}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} ~\vert~
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} ~\vert~
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} ~\vert~
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau ~\vert~
  \hexp : \ztau ~\vert~
  \hlam{x}{\zexp} ~\vert~
  \hap{\zexp}{\hexp} ~\vert~
  \hap{\hexp}{\zexp} ~\vert~
  \hadd{\zexp}{\hexp} ~\vert~
  \hadd{\hexp}{\zexp} ~\vert~
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single cursor.}
\label{fig:zexp-syntax}
\end{figure}

@@

The rules for relative movement within Z-types are given below and should be self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }{
      \aMove{\dChild}
    }{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{{\htau_2}}
    }{
      \aMove{\dNext}
    }{
      {\tarr{\htau_1}{\zwsel{\htau_2}}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dPrev}
    }{
      {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
    }
  }
\end{equation}
% \begin{equation}
% \inferrule{
%   \performTyp{
%     \ztau
%   }{
%     \aMove{\delta}
%   }{
%     \ztau'
%   }
% }{
%   \performTyp{
%     \tarr{\ztau}{\htau}
%   }{
%     \aMove{\delta}
%   }{
%     \tarr{\ztau'}{\htau}
%   }
% }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{
%       \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \ztau'
%     }
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aMove{\delta}
%     }{
%       \tarr{\htau}{\ztau}
%     }
%   }
% \end{equation}
\end{subequations}
% The final two rules above recurse into the zipper structure.

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
\end{subequations}
\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

Finally, the $\aConstruct{\fplus}$ action constructs a plus expression with the focused expression as its first argument. If the focused expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
\begin{equation}
  \inferrule{
    \tcompat{\htau}{\tnum}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fplus}}
      {\hadd{\hexp}{\zwsel{\hehole}}}
      {\tnum}
  }
\end{equation}

@@

If the focused non-empty hole appears in synthetic position, then it can always be finished:
\begin{subequations}
  \begin{equation}
    \label{r:finishana}
  \inferrule{
    \hsyn{\hGamma}{\hexp}{\htau'}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\tehole}
      {\aFinish}
      {\zwsel{\hexp}}
      {\htau'}
  }
\end{equation}

If the focused non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the wrapped expression is consistent with the type the hole is being analyzed against. This amounts to analyzing those contents against the provided type (by subsumption):
\begin{equation}
  \inferrule{
    \hana{\hGamma}{\hexp}{\htau}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\htau}
      {\aFinish}
      {\zwsel{\hexp}}
  }
\end{equation}
\end{subequations}

@@

\end{subequations}

\section{Mechanization}
\label{sec:mech}\label{sec:mt}
\input{agda}

\section{Implementation}
\label{sec:impl}
\input{implementation}

\section{Related Work}\label{sec:rw}
%\subsection{Structure Editors}

Structured editing has been recognized as a way to avoid the possibility of syntax errors for decades.  An early example is the
The Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981}, first published in 1981.
The synthesizer generator~\cite{Reps:1984:SG:390010.808247} allows the user to create an attribute-grammar specification that then can be used to generate a structured editor.
CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language. Barista \cite{ko_barista:_2006} is a modern take on the same basic concept.%These early systems were developed  of the systems are rooted in the type-theoretic tradition.

Novice programmers have been a common target for structure editors. For example,
GNOME\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates.
Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.
Touchdevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students.
Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students. These are largely drag-and-drop user interfaces with a limited action model and an unclear semantics.

Not all structure editors are for educational purposes. For example,
mbeddr \cite{voelter_mbeddr:_2012} is an extensible C-based Programming Language and IDE (nominally, for programming embedded systems.)
mbeddr is build on top of the commercial JetBrains MPS framework for constructing structure editors.
Another popular approach is to bring elements of structured editing into a traditional editor.
Codelets \cite{oney_codelets:_2012} uses structured editing to add interactive documentation and examples in an editor. 
Our previous work on Graphite~\cite{Omar:2012:ACC:2337223.2337324} allows developers to associate structured editing interfaces called  \emph{palettes} with types. Graphite is integrated into a text-based program editor (Eclipse.)

Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g. Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes. These  systems are also, to our knowledge, not formally well-defined but rather exist only as part of system implementations.

Perhaps the systems most similar in spirit to Hazelnut are Lamdu~\cite{lamdu} and Unison~\cite{unison}. Like Hazelnut, these are both statically typed functional language editors. In both cases, the language is similar to Haskell. In Lamdu, the editor uses structure editing to enable Live Programming, where the code is always being executed as it is being written.

Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified. For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model. Unison is also a rather large body of Haskell code, though its action model appears superficially more similar to ours. We maintain what we believe to be a stronger action sensibility invariant than Unison (i.e. in Unison, one must construct expressions from the outside-in.) These systems are rich sources of interesting ideas, however -- there is room enough for many different approaches in this (re-)emerging space.

%Drag-and-drop / for novices: lots of examples, e.g. Alice and others
%
%Contemporary: Lamdu, MPS/Mbeddr, TouchDevelop
%
%Hybrid: Cyrus' active code completion paper

%\subsection{Refactoring Models}
%(Michael, can you fill this section out?)

%\subsection{Formal Editor Models}
%Need to do a search to see what else has been done...

\section{Discussion \& Conclusion}
\label{sec:future}
This paper presented Hazelnut, a type theoretic structure editor calculus. Our aim is to take a principled approach to its design by formally specifying its semantics, providing strong metatheoretic guarantees, mechanizing its semantics and metatheory in Agda and implementing it using the concepts of  functional reactive programming. As of this submission, we have achieved reasonable confidence in the formal system presented above, and have transitioned our focus toward the mechanization and implementation efforts. By the time of presentation, we anticipate having complete or nearly complete versions of these.

>>>>

% !TEX root = hazelnut-popl17.tex

The previous section introduced Hazelnut by example. In this section, we systematically introduce the following  structures:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item \textbf{H-expressions} and \textbf{H-types} (Sec. \ref{sec:holes}), i.e. expressions and types with \emph{holes}. H-types classify H-expressions according to Hazelnut's \textbf{bidirectionally typed static semantics}.
\item \textbf{Z-expressions} and \textbf{Z-types} (Sec. \ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-expressions and H-types, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-expression (resp. Z-type) corresponds to an H-expression (resp. H-type) by \emph{cursor erasure}.
\item \textbf{Actions} (Sec. \ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectionally typed action semantics}. The action semantics enjoys a rich metatheory. Of particular note, the \emph{sensibility theorem} establishes that every edit state is well-typed after cursor erasure.
\end{itemize}

Our overview below omits certain ``uninteresting'' details. The supplement includes the complete collection of rules, in definitional order. These rules, along with the proofs of the metatheorems, have been mechanized using Agda \cite{norell:thesis}, also in the supplement. We will summarize this effort in Sec. \ref{sec:mech}.% The supplement contains the Agda sources and additional technical details.

\subsection{Holes}\label{sec:holes}

@@

\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \tau,\htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & e,\hexp & ::= &
  \hexp : \htau ~\vert~
  x ~\vert~
  \hlam{x}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}

We refer to terms that do not contain subterms of hole form as \emph{complete}. Informally, we will use metavariables $\tau$ and $e$ rather than $\htau$ and $\hexp$ for complete H-types and H-expressions, respectively. Formally, we can derive $\hcomplete{\tau}$ when $\tau$ is a complete H-type, and $\hcomplete{e}$ when $e$ is a complete H-expression. We omit the straightforward definitions of these judgements for concision. The dynamics of Hazelnut, which we need not detail here, is defined only  over complete H-expressions (i.e. we can only ``run'' a complete program, though see Sec. \ref{sec:future}.)\todo{get rid of e/edot distinction}\todo{change this to refer to dynamics section}

@@

\todo{redundant text w/ citation}
The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}. Their system (which was not bidirectionally typed nor an editor model) also needed to define two rules for function application. In general, when a premise requires that a synthesized type be of a particular form, we need a special case where the synthesized hole type is treated instead as if it were the ``holey-est'' type of that form.\footnote{Alternatively, we might add a rule that allows expressions that synthesize hole type to then non-deterministically synthesize any other type, but maintaining determinism is useful in practice, so we avoid this approach.}

\end{subequations}
\subsection{Focus Model}\label{sec:cursors}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} ~\vert~
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} ~\vert~
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} ~\vert~
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau ~\vert~
  \hexp : \ztau ~\vert~
  \hlam{x}{\zexp} ~\vert~
  \hap{\zexp}{\hexp} ~\vert~
  \hap{\hexp}{\zexp} ~\vert~
  \hadd{\zexp}{\hexp} ~\vert~
  \hadd{\hexp}{\zexp} ~\vert~
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single focus.}
\label{fig:zexp-syntax}
\end{figure}

@@

The rules for relative movement within Z-types are given below and should be self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }{
      \aMove{\dChild}
    }{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-parent-arr-right}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-nextSib-arr}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{{\htau_2}}
    }{
      \aMove{\dNext}
    }{
      {\tarr{\htau_1}{\zwsel{\htau_2}}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dPrev}
    }{
      {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
    }
  }
\end{equation}
% \begin{equation}
% \inferrule{
%   \performTyp{
%     \ztau
%   }{
%     \aMove{\delta}
%   }{
%     \ztau'
%   }
% }{
%   \performTyp{
%     \tarr{\ztau}{\htau}
%   }{
%     \aMove{\delta}
%   }{
%     \tarr{\ztau'}{\htau}
%   }
% }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{
%       \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \ztau'
%     }
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aMove{\delta}
%     }{
%       \tarr{\htau}{\ztau}
%     }
%   }
% \end{equation}
\end{subequations}
% The final two rules above recurse into the zipper structure.

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:\todo{remove prevSib}
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}\label{rule:move-parent-asc-right}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
and this for the example\todo{revise}
\begin{equation}\label{r:movefirstchild-lam}
\inferrule{ }{
  \performMove{
    \zwsel{\hlam{x}{\hexp}}
  }{
    \aMove{\dChild}
  }{
    \hlam{x}{\zwsel{\hexp}}
  }
}
\end{equation}

\begin{equation}\label{r:moveparent-ap2}
  \inferrule{ }{
    \performMove{
      \hap{{\hexp_1}}{\zwsel{\hexp_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hap{\hexp_1}{\hexp_2}}
    }
  }
\end{equation}

\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}
\end{subequations}

\paragraph{Reachability}

We give the following theorem to demonstrate that the rules for movement do
indeed capture the process of moving the focus to any editable position
within a term. Intuitively, given two zippered expressions that are the
same up to the erasure of focus, its proof gives a list of composable
actions that, when applied to the first term produces the second.

As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
every action in such a list, it is possible consider moving to an ill-typed
term. Therefore the statement here matches the mutually recursive structure
of the bidirectional typing judgements.

\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
\label{thrm:reach}
Both of the following hold:
\begin{enumerate}
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
   = \zexp'$.

  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
   = \zexp'$.
\end{enumerate}
\end{theorem}

\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

Finally, the $\aConstruct{\fplus}$ action constructs a plus expression with the focused expression as its first argument. If the focused expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
\begin{equation}\label{rule:construct-plus-compat}
  \inferrule{
    \tcompat{\htau}{\tnum}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fplus}}
      {\hadd{\hexp}{\zwsel{\hehole}}}
      {\tnum}
  }
\end{equation}

@@

If the focused non-empty hole appears in synthetic position, then it can always be finished:
\begin{subequations}
  \begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\hexp}{\htau'}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\tehole}
      {\aFinish}
      {\zwsel{\hexp}}
      {\htau'}
  }
\end{equation}

If the focused non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the wrapped expression is consistent with the type the hole is being analyzed against. This amounts to analyzing those contents against the provided type (by subsumption):
\begin{equation}\label{r:finishana}
  \inferrule{
    \hana{\hGamma}{\hexp}{\htau}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\htau}
      {\aFinish}
      {\zwsel{\hexp}}
  }
\end{equation}
\end{subequations}

\paragraph{Finishability}\todo{do we care?}
As with deletion, since the meaning of the finish action does not depend on
the structure of the term being finished other than requiring that it must
be one of the two hole forms, there is no checksum theorem to state for
this action.

@@

\paragraph{Constructibility}\todo{why is this with the zipper rules not the
    construct rules?}

We give the following theorem to demonstrate that the rules for the action
$\aConstruct{\varphi}$ do indeed capture the process of building any term
of the form $\varphi$. Intuitively, its proof gives a list of composable
actions that, when applied one sequentially to an empty hole in focus,
build any well-typed expression.

As with Theorem \ref{thrm:reach}, the statement of this theorem must match
the mutually recursive structure of the bidirectional typing judgements.

\begin{theorem}[Constructability]\todo{why $\htau$ and not $\ztau$?}
\label{thrm:construct}
Both of the following hold:
\begin{enumerate}
  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
      L \in \mathit{list} ~\mathtt{action}$ such that
      $\mathit{iterate}~ L~ \zwsel{\hhole{}} = \zexp'$ and
      $\removeSel{\zexp'} = \hexp$

  \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
      L \in \mathit{list} ~\mathtt{action}$ such that $\mathit{iterate}~
      L~ \zwsel{\hhole{}} = \zexp'$ and $\removeSel{\zexp'} = \hexp$
\end{enumerate}
\end{theorem}

Because actions are defined on the zippered form of expressions, we argue
that we can construct a term that is the same as the target up to the
erasure of focus rather than on the nose. Since any form inside a zippered
term may be in focus, there are linearly many such terms for a given
$\hexp$---Theorem \ref{thrm:construct} can be composed with
Theorem \ref{thrm:reach} to reach any one that might be desired in
particular.
\end{subequations}
